---
title: 'Exploratory Data Analysis'
subtitle: 'Lab 1, Part 2'
author: "Bailey Kuehl, Erik Sambrailo, Jenna Sparks, Roshni Tajnekar"
date: '2023-03-07'
output: 
  pdf_document:
      fig_caption: true
---

```{r load packages, message=FALSE, echo=FALSE, results=FALSE}
library(tidyverse)
library(dbplyr)
library(glue) #f-string like formatting
library(latex2exp) # for writing Latex in plot text
library(tinytex) # to convert to pdf
library(insight) # format table display
library(knitr)

```

```{r load data, echo=FALSE, results=FALSE}
#Loading in the ANES 2020 voter data, as well as files that contain the variables that we have identified as relevant for our analysis. 

#ANES 2020 data
ANES_2020_complete <- read.csv('../anes_timeseries_2020_csv_20220210.csv')
dim(ANES_2020_complete)#size of original dataset

#df of relevant political affiliation variables
pa_vars <- read.csv('../political_affiliation_var_v2.csv')

#df of relevant difficulty voting variables
dif_vars <- read.csv('../difficulty_var_v2.csv')

#relevant voter variable
vote_vars <- c('V202109x')

#variable for unique ID
ID_vars <- c('V200001')

```

```{r reduce dataset, echo=FALSE, results=FALSE}
# Reducing the dataset by the variables (columns) that we care about and re-assigning the column names for each. 

#combine all the relevant variable columns into one vector
var <- c(ID_vars, vote_vars, pa_vars$variable, dif_vars$variable)
#combine names for all columns
names <- c('ID', 'turnout', pa_vars$title, dif_vars$title)

#create new df of relevant columns
rel_voters <- ANES_2020_complete[,var]
#change column names for new df
colnames(rel_voters) <- c(names)
head(rel_voters, 5)
dim(rel_voters)
```

# Research Question

**Did Democratic voters or Republican voters experience more difficulty voting in the 2020 election?**

# Sub-Questions
Below are some of the sub-questions that need to be answered in order to answer the research question. 

## Voter: How are we defining who is a voter?

Based off the codebook, there appears to be one good summary variable, **V202109X**, that combines three survey questions into an analysis of whether an individual voted or not (*details of how that is derived is in programming section of codebook*)

1) V201024: Which one of the following best describes how you voted (early)?
2) V202066: Did voter vote in November 2020 election?
3) V202051: Where are you registered to vote? *(I am not seeing how this variable is relevant or referenced, but including it since it was listed)*

```{r exploring voter data, echo=FALSE, results=FALSE}
#reviewing the turnout variable and reducing to verified turnout.
# 1 = Voted
# 0 = Did not Vote

table(rel_voters$turnout)#returning value distribution

#reducing dataset 
rel_voters <- rel_voters[rel_voters$turnout == 1,]

dim(rel_voters)
```

Using this variable to review our dataset, we find that **6,450** of the original 8,280 rows contain individuals that are believed to have voted. 

## Political Affiliation: How are we defining the political affiliation of individuals?

There are multiple variables that could assist in defining what political party an individual best aligns with. Similar to voter, there is a calculated variable, **V201231x**, which is based on several other variables and might be our best indicator for the political affiliation of an individual.

It is composed of the following survey variables:

1) V201228: Do you think of yourself as Democrat, Republican, or other?
2) V201229: Would you call your affiliation strong?
3) V201230: If you didn't identify, do you lean either direction?

```{r exploring political affiliation, echo=FALSE, results=FALSE}
#reviewing the party_ID variable and removing useless responses. 
# -9 = Refused 
# -8 = Don't know
# 1-3 Democrat affiliation
# 4 = Independent
# 5-7 Republican affiliation

table(rel_voters$pre_party_id_summary)#returning value distribution

#reducing dataset
pa_voters <- rel_voters[!rel_voters$pre_party_id_summary %in% c(-9, -8, 4),]

#creating a new column for binary political affiliation
pa_voters$pol_affil <- pa_voters$pre_party_id_summary

#assigning 1 for Democrat & 2 for Republican
pa_voters$pol_affil[pa_voters$pre_party_id_summary %in% c(1,2,3)] <- 1
pa_voters$pol_affil[pa_voters$pre_party_id_summary %in% c(5,6,7)] <- 2

#checking new column
table(pa_voters$pol_affil)

dim(pa_voters)
```

There are very few individuals (16) that did not provide information regarding affiliation, and 560 individuals that are categorized as Independent *(Democrat/Republican neutral)*. This reduces our number of individuals with some sort of affiliation lean to **5874**. 

### Double Checking Affiliation

#### Do any of the earmarked "nuetral" individuals show signs political affiliation?


To test this we reviewed individuals from the previous test that responded as Independent.  We checked their responses to other variables that might provide evidence of affiliation.  The variables we have selected are as follows:

1) V201018: What political party are you registered with? (pre-vote)
2) V202064: What political party are you registered with? (post-vote)
3) V202429: Which party represents your views best?
4) V202443: Which party do you feel closest to?
5) V202073: Which President did you vote for?

```{r checking independents, echo=FALSE, results=FALSE} 
# check 1: were any of the eliminated "independent" voters showing strong affiliation in other responses?

#creating a supplementary df of the individuals earmarked as independent to review
independent <- rel_voters[rel_voters$pre_party_id_summary == 4,]

#reviewing a few additional variables for independent individuals for political affiliation clues
table(independent$pre_party_reg)
#pre-vote registration: 1 = Dem, 2 = Rep

table(independent$pst_party_reg)
#post-vote registration: 1 = Dem, 2 = Rep

table(independent$pst_party_views)
#party that best represents views: 1 = Dem, 5 = Rep

table(independent$party_close)
#party you feel closest to: 1 = Dem, 3 = Rep

table(independent$pres_vote)
#Presidential vote: 1 = Biden = Dem, 2 = Trump = Rep

#creating a new column that explores the potential affiliation of the neutral individuals. 
pot_dem <- independent[independent$pre_party_reg == 1
                       & independent$pres_vote == 1,]
dim(pot_dem)

#potential independents that could be republicans
pot_rep <- independent[independent$pre_party_reg == 2
                       & independent$pres_vote == 2,]
dim(pot_rep)
```

Only 2 of these five survey questions were well answered by this subset of Independent voters.  The two questions were the pre-vote registration and presidential vote. Reviewing those two questions, of the 560 individuals, there were 23 that answered Democrat to both, and 9 that answered Republican to both. Since this is such a small quantity compared to our current sample size, we will continue to leave these individuals out of our active dataset.  

#### For the individuals we have grouped by political affiliation, do any of them have survey answers that conflict their current affiliation?

We will use and check the same variables as above against the two political affiliations to see how they align. 

```{r double checking affiliaton, echo=FALSE, results=FALSE} 

#creating temp df of Democrat voters to test
dem_voters <- pa_voters[pa_voters$pol_affil == 1,]

dim(dem_voters)

#creating temp df of Republican voters to test
rep_voters <- pa_voters[pa_voters$pol_affil == 2,]

dim(rep_voters)
```

```{r checking democrats, echo=FALSE, results=FALSE} 
#check 2.1: Are any democrats showing conflicting results for other variables. 

#reviewing a few additional variables 
table(dem_voters$pre_party_reg)
#pre-vote registration: 1 = Dem, 2 = Rep

table(dem_voters$pst_party_reg)
#post-vote registration: 1 = Dem, 2 = Rep

table(dem_voters$pst_party_views)
#party that best represents views: 1 = Dem, 5 = Rep

table(dem_voters$party_close)
#party you feel closest to: 1 = Dem, 3 = Rep

table(dem_voters$pres_vote)
#Presidential vote: 1 = Biden = Dem, 2 = Trump = Rep

#potential democrats that could be republican
pot_dem_to_rep <- dem_voters[dem_voters$pst_party_views == 5
                       & dem_voters$party_close == 3
                       & dem_voters$pres_vote == 2,]
dim(pot_dem_to_rep)
```

Once again some of these variables did not have high response rates, so we are choosing to focus on the three questions. Below are the three questions and the number of Democrat individuals that responded in favor of Republican for each:

1) V202429: Which party represents your views best? **65 Republican responses**
2) V202443: Which party do you feel closest to? **54 Republican responses**
3) V202073: Which President did you vote for? **90 Republican responses**

Of the 3,160 Democrat individuals, **24** provided Republican responses to all three of these questions.  Because this conflicts with how we currently have them grouped, we will be removing them from the dataset. 

```{r checking republicans, echo=FALSE, results=FALSE} 
#check 2.2: Are any republicans showing conflicting results for other variables. 

#reviewing a few additional variables 
table(rep_voters$pre_party_reg)
#pre-vote registration: 1 = Dem, 2 = Rep

table(rep_voters$pst_party_reg)
#post-vote registration: 1 = Dem, 2 = Rep

table(rep_voters$pst_party_views)
#party that best represents views: 1 = Dem, 5 = Rep

table(rep_voters$party_close)
#party you feel closest to: 1 = Dem, 3 = Rep

table(rep_voters$pres_vote)
#Presidential vote: 1 = Biden = Dem, 2 = Trump = Rep

#potential republicans that could be democrats
pot_rep_to_dem <- rep_voters[rep_voters$pst_party_views == 1
                       & rep_voters$party_close == 1
                       & rep_voters$pres_vote == 1,]
dim(pot_rep_to_dem)
```

In our initial Republican grouping, below are the individuals that provided Democrat responses.

1) V202429: Which party represents your views best? **67 Democrat responses**
2) V202443: Which party do you feel closest to? **80 Democrat responses**
3) V202073: Which President did you vote for? **264 Republican responses**

Of the 2,714 Republican individuals, **37** provided Democrat responses to all three of the above questions. We will be removing these from our dataset as well. 

```{r removing questionable affiliation, echo=FALSE, results=FALSE} 
#removing individuals from exercises above from our master dataset, 'pa_voters'. 
#by removing variables 'pot_rep_to_dem' & 'pot_dem_to_rep'

#combining the two df into one
sus_affil <- rbind(pot_dem_to_rep, pot_rep_to_dem)
dim(sus_affil)

#re-assigning master dataset to exclude ID's found in df of suspect affiliation
pa_voters <- pa_voters[!pa_voters$ID %in% sus_affil$ID,]
dim(pa_voters)
dim(pa_voters[pa_voters$pol_affil == 2,])
```
**This leaves us with 5,813 total individuals, 3,136 grouped as Democrat and 2,677 grouped as Republican**

## Difficulty Voting: How are we defining difficulty voting?

```{r review current data, echo=FALSE, results=FALSE, message=FALSE}
#Reviewing the variables specific to difficulty
head(pa_voters[,dif_vars$title],10)

#new vector of just the variables relevant to "types" of difficulties
diff_types <- dif_vars[!dif_vars$title %in% c('pst_likert_diff','diff_none'),]

#creating a new column for counting number of difficulties experienced. 
#assigning 0 to rows with no difficulties
pa_voters$diff_count[pa_voters$diff_none == 1] <- 0

#summing the qty of all types of difficulties experienced
#creates negative numbers where responses included negative-coded response
pa_voters$diff_count <- rowSums(pa_voters[, diff_types$title])

#checking distribution of difficulty quantities
table(pa_voters$diff_count)
 
#checking difficulty values for all negative values
pa_voters[pa_voters$diff_count < 0,]
#above confirms that all individuals returning a negative sum for difficulty count, had invalid responses for each of the difficulty types, so these rows hold no value

#tables of difficulty counts
table(pa_voters[pa_voters$pol_affil == 2
               &pa_voters$diff_count >= 0,
               ]$diff_count)

table(pa_voters[pa_voters$pol_affil == 1
               &pa_voters$diff_count >= 0,
               ]$diff_count)

#initial rough plot at a histogram of the difficulty counts for the two parties
hist(pa_voters[pa_voters$pol_affil == 2
               &pa_voters$diff_count > 0,
               ]$diff_count,
     xlab='Qty of Difficulties Experienced',
     ylab='Qty of Voters',
     main='Republican Difficulties Voting')

hist(pa_voters[pa_voters$pol_affil == 1
               &pa_voters$diff_count > 0,
               ]$diff_count,
     xlab='Qty of Difficulties Experienced',
     ylab='Qty of Voters',
     main='Democrat Difficulties Voting')

```
